Draft Article

### 2019-12-23 article.tex
Complete draft of aerial espionage

\subsection{Inventing the spy plane}
While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow. What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict. What they needed was a spy plane.

For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response. To build such a aircraft, Eisenhower approved Project Aquatone---what became the U-2. From its conception, the U-2 was purpose-built to be read as a civilian aircraft. Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force. \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments. \enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[p.~60. The original source for this quote is an \emph{OSA History} that requires codeword clearance. It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to \enquote{truthfully deny,} in the phrasing of the CIA, that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}

The steps Eisenhower took to ensure the U-2 civilian's character went beyond mere deception. If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient. The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important. The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.

At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology. They counted as espionage because they were espionage! But Eisenhower's precautions were based in the reality of the United State's previous experience with reconnaissance flights. He had already been testing the limits of what the Soviet Union would be willing to accept. No finite list of attributes could ensure that the Russians wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development. But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russian interpret it that way. The U-2 simply had to be a spy plane, in a manner that is both instantly recognizable and ontologically undefinable.

These precautions had the intended effect, even though at first it seemed like the Soviets were going to extract significant concessions. A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation'' and Khrushchev played it up.\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to their cover story, Khrushchev gleefully debunked it by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}

An yet few real consequences followed. No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important. From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin. It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F. Kennedy. If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment. With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.

\subsection{Evaluating the consequences of aerial espionage}
The U-2 Incident is often cited as a seminal failure of the Eisenhower administration. In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible. I disagree. Instead, the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights. There are a number of reasonable explanations for why the U-2 shootdown escalated. Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory. Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia. See: \cite[p.~555]{eisenhower_waging_1965}. Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin. See: \cite[p.~328]{kistiakowsky_scientist_1976}.} What would a major punishment have looked like? One proportional response Khrushchev could have considered was to force the issue on regional air bases. Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights. This likely would have involved some sort of trade, but it is not an unreasonable hypothetical. The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey. Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis. By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights. That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.

Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations. Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy. I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.

Consider the period of time over which these flights took place. If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960. The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident. During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before. Both sides were so willing to contain the issue that many of the airmen never saw their families again.

@@

% As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it. Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot. While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare. Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could. Just as with traditional espionage, the countermeasures did not punish the attempt.

@@

% The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use. Eisenhower's insistence on having the CIA manage Project Aquatone had the intended effect: the spy plane was, in story as well as fact, a civilian intelligence operation.

@@

Both heads of state made clear in their writings that they considered Powers to be spy. On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S. government, and they made no attempt to conceal it. They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War. Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.

Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied. \enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[p.~551]{eisenhower_waging_1965}

The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught. No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}

>>>>

\subsection{Inventing the spy plane}
While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just simply fly a B-52 over Moscow. What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict. What they needed was a spy plane.

For the spy plane to serve its intended purpose as a minimally provocative agent of espionage two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their associated response. To build such an aircraft, Eisenhower approved Project Aquatone---the codename for the U-2. From its conception, the U-2 was purpose-built to be read as a civilian aircraft. Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force. \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments. \enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[p.~60. The original source for this quote is an \emph{OSA History} that requires codeword clearance. It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to \enquote{truthfully deny,} in the phrasing of the CIA, that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}

The steps Eisenhower took to ensure the U-2 civilian's character went beyond mere deception. If his only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient. The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important. The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.

At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology. They counted as espionage because they were espionage! But Eisenhower's precautions were based in the reality of the United State's previous experience with reconnaissance flights. He had already been testing the limits of what the Soviet Union would be willing to accept; if he did everything possible to signal that this plane functioned like a spy, perhaps it would accept this new development as well. No finite list of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development. But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russians interpret it that way as well. The U-2 simply had to be a spy plane, in a manner that is both instantly recognizable and ontologically undefinable.

These precautions had the intended effect, even though it initially seemed like the Soviets were going to extract significant concessions. A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation'' and Khrushchev played it up.\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to their cover story, Khrushchev gleefully debunked it by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}

And yet few real consequences followed. No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important. From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin. It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F. Kennedy. If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment. With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.

Nor did the U-2 Incident end all reconnaissance flights. As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it. Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot. While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare. Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could. Just as with traditional espionage, the countermeasures did not punish the attempt.

\subsection{Evaluating the consequences of aerial espionage}
The U-2 Incident is often cited as a seminal failure of the Eisenhower administration. In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible. I disagree. Instead, the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights. There are a number of reasonable explanations for why the U-2 shootdown escalated. Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory. Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia. See: \cite[p.~555]{eisenhower_waging_1965}. Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin. See: \cite[p.~328]{kistiakowsky_scientist_1976}.} What would a major punishment have looked like? One proportional response Khrushchev could have considered was to force the issue on regional air bases. Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights. This likely would have involved some sort of trade, but it is not an unreasonable hypothetical. The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey. \footnote{Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis.} By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights. That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.

Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations. Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy. I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either. Consider the period of time over which these flights took place; if aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960. The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident. During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before. Both sides were so willing to contain the issue that many of the airmen never saw their families again.

@@

@@

@@

Both heads of state made clear in their writings that they considered Powers to be spy. On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S. government, and they made no attempt to conceal it. They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War. Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding the fundamentals of international espionage still applied. \enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[p.~551]{eisenhower_waging_1965}

% Add slight emphasis here that the U-2 was built as a spy plane because it would have a certain reaction, and it worked

Project Aquatone channeled the norms of espionage in the most crucial respect: its strategic value easily outweighed the consequences of being caught. No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
